---
title: "Mini Project 2"
author: "Eden Ravecca"
date: "12/18/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Mini Project using GBIF Prairie Falcon Data, NLCD Land Cover Data, USFS National Forest Climate Change Data, Tidycensus Cenus Data (by county), and TIGER/Line State shapefiles.  

### <span style="color: navy;"> For the mini project I am interested in exploring whether land cover, historical precipitation averages, and human population influence the liklihood of Prairie Falcon occurrence in Western US States. I downloaded Prairie Falcon data from GBIF, Land Cover data from NLCD, precipitation data from USFS National Forest Climate Change Maps, western states boundary shapefiles from TIGER/Line, and census data from Tidycensus.</span>

#### <span style="color: navy;"> Prairie Falcons are endemic to western North America and are typically found in deserts ranging from Southern Canada to Northern Mexico. Prairie Falcons are rarely seen east of the Mississippi River. They are more commonly found in arid climates, landscapes dominated by shrubs and lacking trees, and are a rare sighting in densely populated areas (unlike other falcons such as Peregrines and Kestrels). I am interested in exploring if their occurrence in the west is driven by low annual precipitation (historical average), land cover types, areas with low human population, or some combination of those variables.</span>    

#### <span style="color: navy;">To begin this project, I downloaded the relevant data from free internet sources. Due to the computational demands of using large spatial and temporal scales, I limited the data to only western states (from the Rocky Mountains to the pacific coast) and focused on a single year for falcon occurrence and land cover- 2016. The dependent variable is Prairie Falcon occurrence observations in 2016. The tabular data I am using includes: states (western US), date (year 2016), counties population. The spatial data includes: land cover 2016, climate (historical precipitation average), state polygons, county polygons.</span>    

```{r, load packages, warning=FALSE}
library(tidyverse)
library(sf)
library(terra)
library(tidycensus)
library(tmap)
library(viridis)
library(RColorBrewer)
library(leaflet)
```

```{r, land cover data}
nlcd_data_terr <- terra::rast("/opt/data/MP/Ravecca_Eden_MP/NLCD/nlcd_2016_land_cover_l48_20210604.img") 
nlcd_data_terr2 <- terra::rast("/opt/data/MP/Ravecca_Eden_MP/NLCD/nlcd_2016_land_cover_l48_20210604.img") 
```

```{r, Prairie Falcon data}
falcon_csv <- read.csv("/opt/data/MP/Ravecca_Eden_MP/FalconData/PRFA_Data.csv")
states <- c( "Idaho", "Montana", "Wyoming", "Utah", "Nevada", "Arizona", "New Mexico", 
             "Colorado", "California", "Oregon", "Washington" ) # to filter states of interest at once
west_prfa <- falcon_csv %>% 
  dplyr::filter(., stateProvince %in% states) %>% # keep western states
  dplyr::select(., c(occurrenceID, locality, stateProvince, decimalLatitude, decimalLongitude, eventDate, day, month, year)) %>% # keep relevant columns
  dplyr::filter(., year == "2016") %>% # keep only 2016 observations
  dplyr::rename(., c(state = stateProvince, lat = decimalLatitude, lon = decimalLongitude, date = eventDate, ID = occurrenceID)) # change names to make sense
prfa.sf <- st_as_sf(west_prfa, coords = c("lon", "lat"), crs = 4326) # convert prfa gps points to sf object
prfa <- st_transform(prfa.sf, crs= crs(nlcd_data_terr))
```

```{r, states data}
states_data <- st_read("/opt/data/MP/Ravecca_Eden_MP/StatesPoly/tl_2012_us_state/tl_2012_us_state.shp")
w_states_shp <- states_data %>%
  dplyr::select(., c(STUSPS, NAME, geometry)) %>% # keep relevant columns
  dplyr::filter(., NAME %in% states) %>% # keep western states of interest
  dplyr::rename(., c(States = NAME, State_Abbreviation = STUSPS))
all(st_is_valid(w_states_shp)) # make sure geometries are valid; TRUE
w_states_reproj <-  st_transform(w_states_shp, crs = crs(nlcd_data_terr)) # reproject states polys to match NLCD data
```

```{r, census data}
census_api_key("ccd5de8eb03adae16092bc5e1c6c40cb104f1e97", install = TRUE, overwrite = TRUE)
readRenviron("~/.Renviron")
options(tigris_use_cache = TRUE)
w.census <- tidycensus::get_estimates(geography = "county",
                                      product = "population",
                                      state = c( "ID", "MT", "WY", "UT", "NV", "AZ", "NM", "CO", "OR", "WA", "CA" ), 
                                      year = 2016,
                                      key = key,
                                      geometry = TRUE,
                                      time_series = T) %>%
  st_transform(., st_crs(w_states_reproj)) %>%
  filter(., variable == "POP") %>%
  spread(., variable, value)
```

```{r, population summary}
cp.summary <- w.census %>% 
   group_by(NAME) %>% 
   summarize(., avg_pop = mean(POP)) %>%
   separate(NAME,c("county","state"),sep=",")
```

```{r, climate data}
precip_hist_terr <- terra::rast("/opt/data/MP/Ravecca_Eden_MP/NFCC_data/Ppt_annual_historical/Ppt_annual_historical.tif") 
x <- rast(extent= ext(nlcd_data_terr), crs= crs(nlcd_data_terr), resolution= 771)
ppt_hist_reproj <- terra::project(precip_hist_terr, x) # project to NLCD crs
```

```{r, crop rasters}
w.states.vect <- as(w_states_reproj, "SpatVector") # turn states shapefiles into SpatVector for cropping (in Terra)
nlcd.fullres.crop <- terra::crop(nlcd_data_terr2, w.states.vect) # crop NLCD full res SpatRaster to States poly
ppt.crop <- terra::crop(ppt_hist_reproj, w.states.vect) # crop Precip. SpatRaster to States poly
```

```{r, aggregate NLCD raster, collapse = TRUE}
nlcd.cats <- terra::cats(nlcd_data_terr2, layer=1) # define nlcd categories
nlcd_modeval <- terra::aggregate(nlcd.fullres.crop, fact=26, fun="modal") # aggregate to 780 m resolution and assign modal value from NLCD classes
```

```{r, resample NLCD raster, collapse = TRUE}
m <- rast(extent= ext(w.states.vect), crs= crs(nlcd.fullres.crop), resolution= 771) # blank raster for template
nlcd_771_mode <- terra::resample(nlcd_modeval, m, method= "near", factors=TRUE) # resample to 771 m resolution and assign NLCD value based on "nearest neighbor"
levels(nlcd_771_mode) <- nlcd.cats # nlcd Land Cover classes as levels
names(nlcd_771_mode) <- "NLCD.Land.Cover.Class"
is.factor(nlcd_771_mode)
```

```{r, spatial join}
cp_prfa <- st_join(prfa, cp.summary, by = "state", left=TRUE)
cp_prfa.vect <- as(cp_prfa, "SpatVector") # turn county population & prfa data into SpatVector for rasterize (in Terra)
```

```{r, static maps}
tm_shape(cp.summary) + 
  tm_polygons(col = "avg_pop",  border.col = "gray", legend.show=TRUE, n=8, style= "cont",
              title = "Population by County (million)") + 
  tm_add_legend(type= "line", lwd = 1, lty= 1, col= "dark gray", title="County Boundaries") +
  tm_legend(outside = TRUE) +
  tm_shape(prfa) + 
  tm_dots(col= "blue", size= 0.01, title= "Prairie Falcons", legend.show=TRUE) +
  tm_add_legend(type= "symbol", size = 0.2, shape = 19, col= "blue", title="Prairie Falcons") +
  tm_legend(outside = TRUE) +
  tm_shape(w_states_reproj) +
  tm_polygons(col = NA, border.col = "black", alpha= 0, title= "Western US States") +
  tm_add_legend(type= "line", lwd = 1, lty= 1, col= "black", title="Western US States") +
  tm_legend(outside = TRUE) +
  tm_layout(main.title = "Prairie Falcon Occurrences in Western US", main.title.size = 1) +
  tm_compass(north = 0, text.size = 0.8, show.labels = 1, 
             cardinal.directions = c("N", "E", "S", "W"), lwd = 1,
             position=c("left", "top"))
brew.ygb <- brewer.pal(8, "YlGnBu") # create color palette object
tmap_mode("plot")
tm_shape(ppt.crop) +
  tm_raster("Ppt_annual_historical", palette= brew.ygb, n=8, legend.show=TRUE, 
            title = "Annual Historical Precipitation (mm)") +
  tm_legend(outside = TRUE) +
  tm_shape(prfa) + 
  tm_dots(col= "black", size= 0.01, title= "Prairie Falcons", legend.show=TRUE) +
  tm_add_legend(type= "symbol", size = 0.2, shape = 19, col= "black", title="Prairie Falcons") +
  tm_legend(outside = TRUE) +
  tm_shape(w_states_reproj) +
  tm_polygons(col = NA, border.col = "dark gray", alpha= 0, title= "Western US States") +
  tm_add_legend(type= "line", lwd = 1, lty= 1, col= "dark gray", title="Western US States") +
  tm_legend(outside = TRUE) +
  tm_layout(main.title = "Prairie Falcon Occurrences in Western US", main.title.size = 1) +
  tm_compass(north = 0, text.size = 0.8, show.labels = 1, 
             cardinal.directions = c("N", "E", "S", "W"), lwd = 1,
             position=c("left", "top"))
brew.set3 <- brewer.pal(7, "Set3") # create color palette object
tm_shape(nlcd_771_mode) +
  tm_raster("NLCD.Land.Cover.Class", palette= brew.set3, n=17, style= "cat", legend.show=TRUE, 
            title = "NLCD Land Cover Types") +
  tm_legend(outside = TRUE) +
  tm_shape(prfa) + 
  tm_dots(col= "navy", size= 0.01, title= "Prairie Falcons", legend.show=TRUE) +
  tm_add_legend(type= "symbol", size = 0.2, shape = 19, col= "navy", title="Prairie Falcons") +
  tm_legend(outside = TRUE) +
  tm_shape(w_states_reproj) +
  tm_polygons(col = NA, border.col = "black", alpha= 0, title= "Western US States") +
  tm_add_legend(type= "line", lwd = 1, lty= 1, col= "black", title="Western US States") +
  tm_legend(outside = TRUE) +
  tm_layout(main.title = "Prairie Falcon Occurrences in Western US", main.title.size = 1) +
  tm_compass(north = 0, text.size = 0.8, show.labels = 1, 
             cardinal.directions = c("N", "E", "S", "W"), lwd = 1,
             position=c("left", "top"))
```

_MW: Thanks for making an attempt Eden. I couldn't get this to knit (maybe you couldn't either?) and it doesn't look like you made all of the assigned figures (including the interactive map)._ 